\phantomsection
\addcontentsline{toc}{chapter}{Preface}
\chapter*{Preface}

This thesis is submitted in fulfillment of the PhD programme in
computer science (\textit{Datalogi}) at the University of Copenhagen,
for Troels Henriksen, under the supervision of Cosmin Eugen Oancea and
Fritz Henglein.

\section{Publications}

Of the peer-reviewed papers I have published during my studies, the
following contribute to this thesis:

\begin{quote}
  \fullcite{henriksen2014size}
\end{quote}
\begin{quote}
  \fullcite{Futhark:redomap}
\end{quote}
\begin{quote}
  \fullcite{henriksen2017futhark}
\end{quote}

\noindent While the following do not:

\begin{quote}
  \fullcite{henriksen2014bounds}
\end{quote}
\begin{quote}
  \fullcite{Henriksen:2016:AGT:2975991.2975997}
\end{quote}
\begin{quote}
  \fullcite{Futhark:segredomap}
\end{quote}

\section{Acknowledgements}

Language design and compiler hacking easily becomes a lonely
endeavour.  I would in particular like to thank Cosmin Oancea, Martin
Elsman, and Fritz Henglein for being willing to frequently and
vigorously discuss language design issues.  I am also grateful to all
students who contributed to the design and implementation of Futhark.
In no particular order: Niels G. W. Serup, Chi Pham, Maya Saietz,
Daniel Gavin, Hjalte Abelskov, Mikkel Storgaard Knudsen, Jonathan
Schrøder Holm Hansen, Sune Hellfritzsch, Christian Hansen, Lasse
Halberg Haarbye, Brian Spiegelhauer, William Sprent, René Løwe
Jacobsen, Anna Sofie Kiehn, Henrik Urms, Jonas Brunsgaard, and Rasmus
Wriedt Larsen.  Development of the compiler was aided by the Giuseppe
Bilotta's advice on OpenCL programming, and James Price's work on
\texttt{oclgrind}~\cite{price2015oclgrind}.  I am also highly
appreciative of the work of all who contributed code or feedback to
the Futhark project on
Github\footnote{\url{https://github.com/diku-dk/futhark}}, including
(but not limited to): David Rechnagel Udsen, Charlotte Tortorella,
Samrat Man Singh, \texttt{maccam912}, \texttt{mrakgr}, Pierre Fenoll,
and \texttt{titouanc}

I am also particularly grateful for Cosmin Oancea's choice to devise a
research programme that has allowed me the opportunity to spend three
years setting up GPU drivers on Linux.  And relatedly, I am grateful
to NVIDIA for donating the K40 GPU used for most of the empirical
results in this thesis.

And of course, I am grateful that Lea was able to put up with me for
these three and a half years, even if I still do not understand why.

The cover image depicts the common European hedgedog
(\textit{Erinaceus europaeus}), and is from \textit{Iconographia
  Zoologica}, a 19th century collection of zoological prints.  Like
hedgehogs, functional languages can be much faster than they might
appear.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "thesis"
%%% End:
